A generalization of the Bradley-Terry model for draws in chess

### 2020-09-14 inst/jebo_chess_rebuttal.txt

“A generalization of the Bradley-Terry model for draws in chess”
JEBO-D-20-00916
Referee Report
This paper develops extensions to the Bradley-Terry model appropriate to Bernoulli trials
so that the model can accommodate the chess-specific needs of draws and the first-mover
advantage possessed by White. The author takes this extended model to various sets of
exclusively game-level outcomes and estimates time-invariant player strengths as well as the
importance of draws and first-mover advantage. After estimating common parameters for draws
and first-mover advantage, the author allows for (time-invariant) player-specific parameters on
these margins. The paper then concludes by considering distinct draw parameters for Soviet and
non-Soviet players. Using data from the top chess events of 1962 and 1963, the author
determines that Soviet players draw more frequently against each other than in other gamescenarios.
The paper does a nice job of clearly laying out the various reified “monsters” that extend
the original Bradley-Terry model to accommodate chess. This framework is plausible and
intuitive. I also appreciated the paper’s brevity in making its point. The allowance for playerspecific tendencies to draw or to gain excess-advantage from moving first also strikes me as an
innovation to the literature. Lastly, the paper is well-written and a fun read.
My comments on the rest of the paper are detailed below by decreasing importance.
1) The author never properly explores the paper’s identification strategy. As I read it, the
identifying assumption, besides the functional form assumption, is that player strengths
are constant over the duration of the considered data. This seems highly plausible within
a specific event (e.g., the Candidates tournament of 1963, the Interzonals of 1962), but
much less so with the Anand-Karpov-Kasparov example. The author does not specify the
timespan over which these 365 games were played, but I imagine that the timespan must
exceed 10 years. To the extent that all players have matured to full strength and none
have yet begun to decline, this assumption would be benign, but it should be supported.
Precise estimation will hinge on a large number of games, but this will be in tension with
the assumption of time-invariant player-strengths. The author should explore this tension.
2) The author neglects to motivate the paper beyond its application within the chess world.
The tournament literature’s applications lie primarily, though definitely not exclusively,
within compensation and labor. How could this extension be used within that context?
Where might economists see the first-mover advantage in action? Where might the
extension to allow draws be especially lucrative? Etc.
3) As the author notes, the work of Moul and Nye (2009) is directly on point. The author’s
discussion of this paper, however, is overly sparse and at times misleading. Moul and

Nye use existing observed player strengths (variations of commonly used scores based on
the Elo rating system that goes back to 1960). Moul and Nye then use these strengths
within an ordered probit to estimate parameters that relate back to draw likelihood and
White’s first-mover advantage. The sharp difference between that paper and this is then
that the author is estimating player relative strengths (conditional upon a pool of players)
while Moul and Nye use “observed” strengths.1 The latter approach is far less
parametrically demanding, but leans much more on the assumption of normally
distributed disturbances (basically the differing assumption of functional form). The latter
approach also allows individuals’ player strength to vary across time and is far more
flexible to changes in the composition of the pool of players.
Furthermore, Moul and Nye estimate several versions of their ordered probit without data
on the number of moves (ply-count). This is contrary to the implication of the paper (p. 6,
immediately below Table 2), which suggests that the ply-count data are essential to that
paper’s findings. All of that paper’s core results (significance of draws, first-mover
advantage, a distinct draw outcome for round-robin international tournament games that
involve two Soviets) are originally derived without ply-count data. The ply-count data are
incorporated to explore a potential mechanism by which players in a draw-cartel might
benefit.
4) The paper concludes that the estimates derived from the 1962 Interzonals and the 1963
Candidates strongly suggest that the “draw monster” won games involving two Soviet
players more frequently than games that involved at least one non-Soviet. This is
consistent with a Soviet cartel, but it could also be the result of Soviet-specific styles of
play. Moul and Nye (2009) test the Soviet style-of-play hypothesis by considering
international tournaments when the format changed from all-play-all to knockout and allSoviet tournaments. Both comparisons yield results consistent with a Soviet draw-cartel
in international all-play-all tournaments and rejecting the Soviet style-of-play hypothesis.
By only considering tournaments that were international and all-play-all, the author
forgoes this critical source of identification.
In the cartel example, the author also neglects the advantage that presumably accrued to
Soviet players when they played non-Soviet players in all-play-all international
tournaments. This would suggest another “monster” to be estimated, one that benefits the
Soviet player when playing a non-Soviet opponent. Without such an addition, Soviet
strengths will be overstated.

1

Those relative strengths are highly predictive, with t-statistics of around 20.

### 2020-09-14 inst/jebo_chess_revision2.tex
revision of JEBO paper, ancillary files at ~/Dropbox/draft_papers/

% -*- mode: noweb; noweb-default-code-mode: R-mode; -*-
\documentclass[nojss]{jss}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{units}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% just as usual
\author{Robin K. S. Hankin\\Department of Mathematical Sciences\\Auckland University of Technology}
\title{A generalization of the Bradley-Terry model for draws in chess}

%\VignetteIndexEntry{chess draws}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Robin K. S. Hankin}
\Plaintitle{Chess Draws}
\Shorttitle{Chess Draws}

%% an abstract and keywords
\Abstract{

In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.}

\Keywords{Likelihood functions, method of support, \proglang{R},
  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
  Draws}
\Plainkeywords{Likelihood functions, method of support, R,
  constrained optimization, Bradley-Terry, Chess, Collusion, Chess
  Draws}

%% publication information
%% NOTE: This needs to filled out ONLY IF THE PAPER WAS ACCEPTED.
%% If it was not (yet) accepted, leave them commented.
%% \Volume{13}
%% \Issue{9}
%% \Month{September}
%% \Year{2004}
%% \Submitdate{2004-09-29}
%% \Acceptdate{2004-09-29}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Robin K. S. Hankin\\
  Auckland University of Technology\\
  E-mail: \email{hankin.robin@gmail.com}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/1/31336-5053
%% Fax: +43/1/31336-734

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}

\SweaveOpts{}
\begin{document}

<<echo=FALSE,print=FALSE>>=
ignore <- require(magrittr,quietly=TRUE)
ignore <- require(hyper2,quietly=TRUE)
options("width" = 90)
@

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry model~\citep{bradley1952}
considers~$n$ players and assigns a non-negative number $p_i$ to player
$i$, with $\sum p_i=1$.  Then player $i$ beats player $j$ with
probability $\frac{p_i}{p_i+p_j}$.  This simple system has been
modified and extended many times, for example to account for multi-way
competitions such as {\em MasterChef Australia}~\citep{hankin2017},
and non-competitive situations such as consumer
behaviour~\citep{hankin2010} and mating preference~\citep{west2008}.

@@

\cite{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw and
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.  Further,
\cite{langer2013} point out that rational players in an ongoing
competitive situation modify their strategies depending on whether
they are leading or lagging: the lagging player will adopt a higher
risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will be more likely to
aim for a draw.

@@

\subsection{First-mover advantage} Traditionally White moves first in
chess, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of~\cite{adams2007}---``White
to play and win'' to~\citet{adorjan1988}---``Black is OK!''.

@@

where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system may be
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, we
seek a likelihood function for non-negative $p_1,\ldots,p_n,\draw$
(here \draw\ is the draw monster's strength, fraktur font being used
for reified entities' strengths) with $\draw+\sum p_i=1$.  A suitable
likelihood function would be

@@

where $W_{ij}$ is the number of games won by $i$ when playing $j$,
$L_{ij}$ is games lost and $D_{ij}$ is games drawn.  However, this
approach ignores the extra strength conferred by the first-move
advantage.  We thus consider another entity whose strength \white\ is
added to the white player.  The strength of the white player is thus
the sum of the player's strength and \white.  Recalling that a game
might be a draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
defined on $\white+\draw+\sum p_i=1$, given below in equation
\ref{likelihood_WD}:

@@

\begin{table}
\centering
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+\white\oo}{p_i+p_j+\white+\draw}$ & $\frac{p_i}{p_i+p_j+\white+\draw}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{\draw\oo}{p_i+p_j+\white+\draw}$     & $\frac{\draw}{p_i+p_j+\white+\draw}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+\white+\draw}$   & $\frac{p_j+\white}{p_i+p_j+\white+\draw}$\\
\end{tabular}
\caption{Probabilities of a win, draw, loss for the two players under a generalized
  Bradley-Terry model\label{probabilities}}
\end{table}

@@

where $W_{ij}^\mathrm{white}$ is the number of games won by $i$ when
playing white against $j$, and so on.  Note that the power of the
final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the total
number of games drawn between $i$ and $j$ irrespective of colour.  We
now apply this and related likelihood models to real chess datasets
using the Method of Support~\citep{edwards1972}.  To test any
hypothesis $H$ on likelihood function $\mathcal{L}$ one first
identifies the constraint on hypothesis space corresponding to $H$.
One then finds the unconstrained maximum likelihood
$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
constrained by $H$.  The test statistic is then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

\cite{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \cite{joe1990} uses likelihood methods
to show that this model does not fit data well.  Here we note that
such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \cite{moul2009} consider chess draws and use sophisticated
regression techniques to analyse games in terms of not only the
result, but also the ply-count, data not used in the present analysis.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the
R~\citep{rcore2019} \pkg{hyper2} package~\citep{hankin2017} as a
convenient and unified interface.  Because gradients are available,
convergence to the evaluate is generally rapid and hypothesis-specific
constraints on parameter space are, in general, linear and
straightforward to implement.  The total runtime for all the tests
presented here is less than a few seconds on a desktop PC.

@@

Grandmasters Anand, Karpov, and Kasparov have long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\centering
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
  Anand played white against Karpov a total of $18+20+13=51$ times and
  won 18, drew 20, lost 13}
\end{table}

@@

<<echo=FALSE,print=FALSE,results=hide>>=
maxp(karpov_kasparov_anand)
@

\begin{figure}[t]
\begin{center}
<<ddc,echo=FALSE,fig=TRUE>>=
dotchart(maxp(karpov_kasparov_anand),pch=16,xlim=c(0,0.55))
@
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

@@

<<maxlikekka,echo=FALSE,print=FALSE,results=hide>>=
small <- 1e-4

e <- function(white){
    maxp(
        karpov_kasparov_anand,
        startp=c(1/3-small,1/3-small,1/3-small,small),
        fcm=-cbind(0, 0, 0, 1),fcv= -white,give=TRUE)$value
}
ignore1 <- e(0.001)
ignore2 <- e(0.002)
ignore3 <- e(0.003)
ignore4 <- maxp(karpov_kasparov_anand,give=TRUE)$value
support <- 328.2975-335.6025
@

<<ignore34diff,echo=FALSE,print=FALSE,results=hide>>=
ignore3
ignore4
ignore3-ignore4
@

The result of numerical optimization is about 7.01 units of support,
or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
the two-units-of-support per degree of freedom suggested by
\citeauthor{edwards1972}; we can convincingly reject the hypothesis
that $\white=0$ and infer that playing white confers a significant
advantage.  This is consistent with expectation, but it is now
possible to go significantly beyond mere existence of the effect and
to quantify it, accounting for the differential strengths of the
players.  Taking the MLE above, and considering a game between say
Karpov and Anand, the strengths are Karpov=0.064, Anand=0.151, \white
=0.104 and \draw =0.532.  Thus we would estimate that Karpov playing
white against Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
Karpov plays black).

<<kka_equal_strength, print=FALSE,echo=FALSE>>=
objective <- function(V){
  x <- V[1]  # strength of all three players
  w <- V[2]  # white player advantage
  D <- 1-3*x-w  # Draw monster
  return(  
      -loglik(c(x,x,x,w),karpov_kasparov_anand)  # negative as we minimize
  )
}

jj <-
constrOptim(
    theta=c(0.1,0.4),
    f=objective,
    grad=NULL,
    ui = rbind(diag(2),-diag(2),c(-3,-1)),
    ci = c(0,0,-1,-1,-1)
)

unconstrained <- maxp(karpov_kasparov_anand,give=TRUE)$value
constrained <- jj$value

extra_support <-   unconstrained + constrained  # negative used
pvalue <- pchisq(2*extra_support,df=2,lower.tail=FALSE)
@

@@

\begin{equation}\label{KKA_personalised_drawmonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
%,\white,
%\draw_\mathrm{An},\draw_\mathrm{Kr},\draw_\mathrm{Ks}
%\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + \white\right)^{25}
\left(p_\mathrm{Ks} + \white\right)^{45} 
\left(p_\mathrm{An} + \white\right)^{24}\\ \rule{10mm}{0mm}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{129}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{49}
\left(\draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}
$}}
\end{equation}

\begin{figure}[t]
\begin{center}
<<echo=FALSE,fig=TRUE>>=
dotchart(maxp(kka_3draws),pch=16)
@ 
\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in
  equation  \ref{KKA_personalised_drawmonster}
} 
  \end{center}
\end{figure}

In Figure~\ref{kka_3draws}, we see considerable variability in
proclivity to draw.  Again this may be tested with the Method of
Support, which gives a $p$-value of about 0.039, borderline
significant: an interesting result, given that in this case no draw
monster appears alone anywhere in the likelihood function.

>>>>

\documentclass[review]{elsarticle}

\usepackage{lineno,hyperref}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{units}
\usepackage{xcolor}
\modulolinenumbers[5]

\journal{Journal of Economic Behavior and Organization}

%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%

%% Numbered
%\bibliographystyle{model1-num-names}

%% Numbered without titles
%\bibliographystyle{model1a-num-names}

%% Harvard
\bibliographystyle{model2-names.bst}\biboptions{authoryear}

%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}

%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}

%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}

%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}

%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\begin{frontmatter}

\title{A generalization of the Bradley-Terry model for draws in chess}

%% Group authors per affiliation:
\author{Robin K. S. Hankin\fnref{myfootnote}}
\address{AUT University, 2-14 Wakefield Street, Auckland, New Zealand}

\begin{abstract}
In inference problems where the dataset comprises Bernoulli outcomes
of paired comparisons, the Bradley-Terry model offers a simple and
easily interpreted framework.  However, it does not deal easily with
chess because of the existence of draws, and the white player
advantage.  Here I present a new generalization of Bradley-Terry in
which a chess game is regarded as a three-way competition between the
two players and an entity that wins if the game is drawn.
Bradley-Terry is then further generalized to account for the white
player advantage by positing a second entity whose strength is added
to that of the white player.  These techniques afford insight into
players' strengths, response to playing black or white, and
risk-aversion as manifested by probability of drawing.  The likelihood
functions arising are easily optimised numerically.  I analyse a
number of datasets of chess results, including the infamous 1963 World
Chess Championships, in which Fischer accused three Soviet players of
collusion.  I conclude, on the basis of a dataset that includes only
the scorelines at the event itself, that the Candidates Tournament
(Cura\c{c}ao 1962) exhibits evidence of collusion ($p<10^{-5}$), in
agreement with previous work that uses additional information such as
historical results including number of moves.
\end{abstract}

\begin{keyword}
  Chess draws\sep Bradley-Terry\sep Likelihood \sep Chess Collusion
\end{keyword}

\end{frontmatter}

\linenumbers

\newcommand\headercell[1]{%
   \smash[b]{\begin{tabular}[t]{@{}c@{}} #1 \end{tabular}}}
\newcommand{\draw}{\ensuremath{\mathfrak{D}}}
\newcommand{\cdraw}{\ensuremath{\mathfrak{C}}}
\newcommand{\white}{\ensuremath{\mathfrak{W}}}
\newcommand{\Prob}{\operatorname{Prob}}

@@

Many situations involve pairwise comparisons that amount to a
Bernoulli trial.  The canonical application would be two-way
competitive situations such as tennis in which one player is
identified as the winner.  The Bradley-Terry
model~\citep{bradley1952,turner2012} considers~$n$ players and assigns
a non-negative number $p_i$ to player $i$, with $\sum p_i=1$.  Then
player $i$ beats player $j$ with probability $\frac{p_i}{p_i+p_j}$.
This simple system has been modified and extended many times, for
example to account for multi-way competitions such as {\em MasterChef
  Australia}~\citep{hankin2017}, and non-competitive situations such
as consumer behaviour~\citep{hankin2010} and mating
preference~\citep{west2008}.

@@

\citet{moul2009} conceptualise a player as being able to adopt a
strategy on a spectrum ranging from conservative to aggressive, thus
being able to exert some control over the probability of a draw and
suggest that risk-averse players may pursue less vigorous strategies,
trading expected payout for reduced variance.  Further,
\citet{langer2013} point out that rational players in an ongoing
competitive situation modify their strategies depending on whether
they are leading or lagging: the lagging player will adopt a higher
risk strategy (`nothing to lose'), while the leading player is
incentivised to protect the lead by lowering the risk of losing.  In a
chess context, this means the leading player will be more likely to
aim for a draw.

@@

\subsection{First-mover advantage} Traditionally White moves first in
chess, and this feature introduces an asymmetry into play.
Historically, we see about 1.35 white wins for every black win.
However, interpretations of this statistic vary and the issue is hotly
debated with opinions ranging from that of \citet{adams2007}---``White
to play and win'' to \citet{adorjan1988}---``Black is OK!''.

@@

\noindent where $W_{ij}$ is the number of times $i$ beats $j$ and $L_{ij}$ is
the number of times $i$ loses to $j$.  Here I present a new
generalization of Bradley-Terry in which a chess game is regarded as a
three-way competition between the two players and a third entity (a
``draw monster'') that wins if the game is drawn.  The system may be
further modified to account for first-move advantage by positing an
entity whose strength is added to that of the white player.  This
methodology affords insight into players' strengths, response to
playing black or white, and risk-aversion as manifested by probability
of drawing.  If we consider a chess game to be a three-way trial, we
seek a likelihood function for non-negative $p_1,\ldots,p_n,\draw$
(here \draw\ is the draw monster's strength, fraktur font being used
for reified entities' strengths) with $\draw+\sum p_i=1$.  A suitable
likelihood function would be

@@

\noindent where $D_{ij}$ is games drawn.  However, this approach
ignores the extra strength conferred by the first-move advantage.  We
thus consider another entity whose strength \white\ is added to that
of the white player.  The strength of the white player is thus the sum
of the player's strength and \white.  Recalling that a game might be a
draw, we obtain the probabilities as given in
Table~\ref{probabilities}.  The corresponding likelihood function is
defined on $\white+\draw+\sum p_i=1$, given below in equation
\ref{likelihood_WD}:

@@

\begin{table}
  \caption{Probabilities \label{probabilities}of a win, draw, loss for the two players under a generalized Bradley-Terry model}
\centering
\fbox{%
\begin{tabular}{l|cc}\\
  &$i$ plays white & $j$ plays white\\ \hline
$\Prob{(\mbox{$i$ wins})}$  & $\frac{p_i+\white\oo}{p_i+p_j+\white+\draw}$ & $\frac{p_i}{p_i+p_j+\white+\draw}$  \\
$\Prob{(\mbox{draw})}$      & $\frac{\draw\oo}{p_i+p_j+\white+\draw}$     & $\frac{\draw}{p_i+p_j+\white+\draw}$    \\
$\Prob{(\mbox{$i$ loses})}$ & $\frac{p_j\oo}{p_i+p_j+\white+\draw}$   & $\frac{p_j+\white}{p_i+p_j+\white+\draw}$\\
\end{tabular}}
\end{table}

@@

\noindent where $W_{ij}^\mathrm{white}$ is the number of games won by
$i$ when playing white against $j$, and so on.  Note that the power of
the final term, $D_{ij}^\mathrm{white}+D_{ij}^\mathrm{black}$ is the
total number of games drawn between $i$ and $j$ irrespective of
colour.  We now apply this and related likelihood models to real chess
datasets using the Method of Support~\citep{edwards1972}.  To test any
hypothesis $H$ on likelihood function $\mathcal{L}$ one first
identifies the constraint on hypothesis space corresponding to $H$.
One then finds the unconstrained maximum likelihood
$\mathcal{L}_\mathrm{free}$ and compares it with the maximum
likelihood $\mathcal{L}_\mathrm{cons}$ when the optimization is
constrained by $H$.  The test statistic is then the likelihood ratio
$\mathcal{L}_\mathrm{free}/\mathcal{L}_\mathrm{cons}$ or, equivalently
the support $\mathcal{S}=\log\mathcal{R}$.  Wilks's theorem states
that the asymptotic distribution of $2\mathcal{S}$ is $\chi^2_n$ where
$n$ is the number of degrees of freedom implemented by $H$.

@@

\citet{davidson1970} uses a threshhold approach to interpret draws.
From a chess draw, he infers that the performance of the players is
identical, within a threshhold.  Using a logistic expression for the
difference in players' performance he uses, in current notation,
$\draw=\nu\sqrt{p_ip_j}$.  However, \citet{joe1990} uses likelihood
methods to show that this model does not fit data well.  Here we note
that such a model automatically give stronger players a higher draw
potential, and also note that it is difficult to account for
first-player advantage in this framework.  Such an approach is
numerically challenging in the current framework as the unit-sum
constraint becomes nonlinear, making likelihood maximization
difficult.  \citet{moul2009} consider chess draws and use
sophisticated regression techniques to analyse games in terms of not
only the result, but also the ply-count, data not used in the present
analysis.

@@

The Method of Support \citep{edwards1972} involves only numerical
optimization (not integration which is computationally expensive).
The analysis presented here uses the framework of the
R~\citep{rcore2019} {\tt hyper2} package~\citep{hankin2017} as a
convenient and unified interface.  Because gradients are available,
convergence to the evaluate is generally rapid; hypothesis-specific
constraints on parameter space are, in general, linear and
straightforward to implement.  The total runtime for all the tests
presented here is less than a few seconds on a desktop PC.

@@

Grandmasters Anand, Karpov, and Kasparov have a long-standing rivalry
and have played many games amongst themselves.  A summary may be found
in Table~\ref{kka}.  Equation~\ref{likelihood_WD} becomes

\begin{table}
\caption{Results of 365 chess games \label{kka} between Grandmaster
  Anand, Grandmaster Karpov, and Grandmaster Kasparov.  For example,
  Anand played white against Karpov a total of $18+20+5=43$ times and
  won 18, drew 20, lost 5}
\centering
\fbox{%
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays White,}\\ \small{wins}} & \multicolumn{3}{c@{}}{\small{Plays black, loses}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  18 &   6  \\
  Kr &  7 &  -  &  18  \\
  Ks & 15 &  30 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{draws}} & \multicolumn{3}{c@{}}{\small{Plays black, draws}}\\
 \cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &  20 &  20  \\
  Kr & 29 &  -  &  72  \\
  Ks & 26 &  57 &  -   \\
\end{tabular}\quad
\begin{tabular}{@{} *{4}{c} @{}}
\headercell{\small{Plays W,}\\ \small{loses}} & \multicolumn{3}{c@{}}{\small{Plays black, wins}}\\
\cmidrule(l){2-4}
& An &  Kr & Ks        \\ 
\midrule
  An & -  &   5 &   11  \\
  Kr & 13 &  -  &    9  \\
  Ks &  2 &   7 &  -   \\
\end{tabular}}
\end{table}

@@

\begin{figure}[t]
\begin{center}
\makebox{\includegraphics{collusion_chess_draws-ddc.pdf}}
\caption{Maximum likelihood estimate for the strengths of the 3 Grandmasters in Table~\ref{kka} 
\label{maxp_karpov_kasparov_anand} under a likelihood defined as Equation~\ref{likelihood_WD}}
  \end{center}
\end{figure}

@@

The result of numerical optimization is about 7.01 units of support,
or a likelihood ratio of $e^{7.01}\simeq 1107.7$: that is, well over
the two-units-of-support per degree of freedom suggested by Edwards;
we can convincingly reject the hypothesis that $\white=0$ and infer
that playing white confers a significant advantage,
consistent with expectation.  It is now possible to go beyond
merely establishing the existence of the effect, and to quantify it
accounting for the differential strengths of the players.  Taking the
MLE above, and considering a game between say Karpov and Anand, the
strengths are Karpov=0.064, Anand=0.151, \white =0.104 and \draw
=0.532.  Thus we would estimate that Karpov playing white against
Kasparov would win with probability
$\frac{0.064+0.104}{0.064+0.104+0.148+0.532}\simeq 0.198$, draw with
probability $\frac{532}{64+104+148+532}\simeq 0.625$, and lose with
probability $\simeq 0.174$ (compare $+0.075\ {=}0.627\ {-}0.297$ if
Karpov plays black).

@@

\begin{equation}\label{KKA_personalised_drawmonster}
%\mathcal{L}\left(
%p_\mathrm{An},p_\mathrm{Kr},p_\mathrm{Ks}
%,\white,
%\draw_\mathrm{An},\draw_\mathrm{Kr},\draw_\mathrm{Ks}
%\right)=
\frac{
\parbox{3in}{$
p_\mathrm{An}^{15}
p_\mathrm{Kr}^{12}
p_\mathrm{Ks}^{20}
\left(p_\mathrm{Kr} + \white\right)^{25}
\left(p_\mathrm{Ks} + \white\right)^{45} 
\left(p_\mathrm{An} + \white\right)^{24}\\ \rule{10mm}{0mm}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{129}
\left(\draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{49}
\left(\draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{46}
$}}{\parbox{3in}{$
\left(p_\mathrm{Kr} + p_\mathrm{Ks} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{Ks}\right)^{193}\\ \rule{10mm}{0mm}
\left(p_\mathrm{Kr} + p_\mathrm{An} + \white + \draw_\mathrm{Kr} + \draw_\mathrm{An}\right)^{92}\\ \rule{20mm}{0mm}
\left(p_\mathrm{Ks} + p_\mathrm{An} + \white + \draw_\mathrm{Ks} + \draw_\mathrm{An}\right)^{80}$}}
\end{equation}

\begin{figure}[t]
\begin{center}
\makebox{\includegraphics{collusion_chess_draws-007.pdf}}
\caption{Maximum likelihood estimate for \label{kka_3draws} the support function given in equation  \ref{KKA_personalised_drawmonster}} 
  \end{center}
\end{figure}

Figure~\ref{kka_3draws} shows the maximum likelihood estimate for the
terms of equation~\ref{KKA_personalised_drawmonster}.  We see
considerable variability in proclivity to draw.  A null of equal
proclivity to draw may be tested with the Method of Support, which
gives a $p$-value of about 0.039, borderline significant: an
interesting result, given that in this case no draw monster appears
alone anywhere in the likelihood function.
